---
title: 'Introduction to Linear Regression: R Notebook'
output:
  html_document: default
  html_notebook: default
  pdf_document: default
---

This notebook demonstrates how to plot data and fit a linear regression.

# Linear Regression

**Step 1**: Download data for the "Introduction to Statistical Learning" (*you may need to do this manually on non-linux operating systems*)
```{bash}
cd /tmp
wget http://www-bcf.usc.edu/~gareth/ISL/Advertising.csv
```

**Step 2**: Load the dataset from the CSV
```{r}
ads <- read.csv("/tmp/Advertising.csv")
```


**Step 3**: Fit a linear regression model
```{r}
lm.fit = lm(Sales ~ TV, data=ads)
summary(lm.fit)
```


**Step 4**: Plot the fit
```{r}
#pdf("/tmp/sales_tv_reg.pdf",7,5)
plot(ads$TV, ads$Sales,col='red',pch=20,xlab = "TV", ylab = "Sales")
abline(lm.fit)
#dev.off()
```

**Step 5:** Plot advanced properties of the fit:
```{r}
plot(lm.fit)
```

# Residuals for Advertising Sales

Now lets look at the regression of the sales as a function of the TV, Radio, and Newspaper advertising.

## TV

```{r}
#pdf("/tmp/sales_tv.pdf", 7,5)
plot(ads$TV, ads$Sales,col='red',pch=20,xlab = "TV", ylab = "Sales")
lm.fit <- lm(Sales ~ TV, data=ads)
abline(lm.fit)
#dev.off()
```

```{r}
sprintf('R-squared: %f', summary(lm.fit)$r.squared)
sprintf('Correlation^2: %f', cor(ads$Sales,ads$TV)^2)
```

## Radio

```{r}
#pdf("/tmp/sales_radio.pdf", 7,5)
plot(ads$Radio, ads$Sales,col='red',pch=20,xlab = "Radio", ylab = "Sales")
lm.fit <- lm(Sales ~ Radio, data=ads)
abline(lm.fit)
#dev.off()
```

```{r}
sprintf('R-squared: %f', summary(lm.fit)$r.squared)
sprintf('Correlation^2: %f', cor(ads$Sales,ads$Radio)^2)
```

## Newspaper

```{r}
#pdf("/tmp/sales_newspaper.pdf", 7,5)
plot(ads$Newspaper, ads$Sales,col='red',pch=20,xlab = "Newspaper", ylab = "Sales")
lm.fit <- lm(Sales ~ Newspaper, data=ads)
abline(lm.fit)
#dev.off()
```

```{r}
sprintf('R-squared: %f', summary(lm.fit)$r.squared)
sprintf('Correlation^2: %f', cor(ads$Sales,ads$Newspaper)^2)
```